---
title: "Episode 2:The 1992 Parliamentary Results"
author: "Mary Jonah"
date: '2018-08-18'
slug: episode-2-the-1992-parliamentary-results
categories: ["elections", "ghana"]
tags: []
---

Welcome to the second episode on my journey to get data on Ghana's elections since 1992.In the [first](http://www.maryjonah.me/2018/08/episode-1-the-1992-presidential-results-rmd/) post, we cleaned the 1992 presidential elections,click [here](https://github.com/maryjonah/maryjonah.github.io/blob/master/ghana_elections/1992/presidential_results_1992.csv) for the data. Today we are going to walk through the process of cleaning the 1992 parliamentary results. 

General elections are held every 4 years in Ghana, during the elections there are 2 positions voters are to fill:

1. **Presidential**: The person with the highest vote becomes the president of Ghana. He/She is an individual who represents a political party and is selected by delegates who represent a section of people in that political party. 
2. **Parliamentary**: Unlike presidential candidates,only people from a constituency (there are multiple constituencies in a region) vote for parliamentary aspirants.

Now back to the main course; *parliamentary aspirant name with hot spicy constituency differentiated by party name and for dessert the region the constituency is located in*.

## Importing Libraries and Data Extraction
The source of this data was [Wikipedia](https://en.wikipedia.org/wiki/MPs_elected_in_the_Ghanaian_parliamentary_election,_1992). On the page I chose the data on the main table. My favourite R library to scrape data on the web is ***rvest***. ***tidyverse*** is another package I love to work with. It is an amazing collection of other R packages *(dplyr, tidyr, ggplot2, forcats, purrr)* which make various aspects of data analysis very easy; data manipulation, reading data files, string manipulation.
Remeber that these 2 packages do not come with the base R hence they need to be installed via the command
`install.packages('packageNameHere')`

```{r message=FALSE, warning=FALSE}
library(rvest)
library(tidyverse)

# get data from Wikipedia and convert to table format
html_Parliament <- read_html("https://en.wikipedia.org/wiki/MPs_elected_in_the_Ghanaian_parliamentary_election,_1992")
html_MP <- html_table(html_nodes(html_Parliament, ".wikitable")[[2]],
           fill = TRUE)

# print the first few rows of the dataframe
head(html_MP)
```
## Extracting Regional Information
html_MP currently is a dataframe with 220 rows with 6 columns. Take a look at the data [here](https://github.com/maryjonah/maryjonah.github.io/blob/master/ghana_elections/1992/raw_data_1992_parliamentary_results.csv).

From the link, we can see that the parliamentary results are stored as follows:

**
1. The region name with the number of seats: repeated across all 6 columns.
2. The column headings: Constituency, Elected MP, Elected Party and 3 NA columns.
3. The rest are rows each representing a parliamentary constituency result in the region **1**.
**

Remember these are repeated for each constituency and region. The question now is **How do you clean the dataframe and get each constituency/region combination?**.
I was able to answer the question above. My method may be clumsy but it worked. 

A dataframe containing all the rows with the word **Region** was created from **html_MP**. You may not find the use of this dataframe but trust me it will come in handy later.
```{r message=FALSE, warning=FALSE}
# dataframe of rows with the word "Region"
match_Region <- html_MP[grepl("Region",html_MP$X1), ]
match_Region

# get the rows(in numeric format) from the dataframe above
regionNum <- rownames(match_Region)
regionNum <- as.numeric(regionNum)
regionNum
```

### Get names of Regions
We will want to group each constituency result by the region they are found in. So a variable **regionNames** is created from the **html_MP** and it will contain the names of the regions.
```{r message=FALSE, warning=FALSE}
regionNames <- match_Region$X1
regionNames <- gsub(" \\-.+\\]", "", regionNames)
regionNames <- gsub("Region", "", regionNames)

# remove empty rows on the right
regionNames <- trimws(regionNames, "right")
regionNames
```

### Create empty dataframes
Next we create 10 empty dataframes. We're going to name the dataframes in the format **"dframe_"+regionNames**. These dataframes are to contain the constituency results for each region.
```{r message=FALSE, warning=FALSE}
x=1
while(x <= 10){
    colsx <- data.frame(
        Constituency = factor(),
        `Elected MP` = factor(),
        `Elected Party` = factor(),
        col4 = factor(),
        col5 = factor(),
        col6 = factor(),
        stringsAsFactors = TRUE
    )
    assign(paste("dframe_", regionNames[x], sep=""), colsx)
    x = x+1
}
```

Next we store the created dataframes into a list; **allRegions**. The dataframes were intentionally created with the word **dframe_** starting each of them. We will store all variables starting with *dframe_* as a list. A combination of `ls()` and `grepl()` is used to get the variables.
```{r message=FALSE, warning=FALSE}
# show all variables beginning with dframe_ in the current session.
grepl("^dframe_", ls())
allRegions <- list(ls()[grepl("^dframe_",ls())])

# print the content of allRegions
allRegions
```

## Extract Rows into Dataframes
Now the read deal. All we have done till now is to get the ingredients. It is time to cook. My intention is to extract all the rows between 2 regionNum values ; .i.e. 1st to 36th rows into a dataframe, 36th to 59th row into another dataframe ... 179th and 200th row into a dataframe. One might understand the use of this method when one takes a look at the raw dataframe [**html_MP**](https://github.com/maryjonah/maryjonah.github.io/blob/master/ghana_elections/1992/raw_data_1992_parliamentary_results.csv) once again. The rows corresponding to each 2 row numbers corresponds to a parliamentary results for a region.
To track the row numbers, 2 numeric vectors are created **prevNum and currentNum**
```{r message=FALSE, warning=FALSE}
prevNum <- numeric()
currentNum <- numeric()

# collect each group of rows and assign to dataframe in list
for(i in 1:length(regionNum)){
    if(i == 1){
        prevNum = 200
        currentNum = 220
        allRegions[[1]] <- html_MP[prevNum:currentNum, ]
    }
    else{
        prevNum <- regionNum[i-1]
        currentNum <- regionNum[i]
        
        # assign each element (dataframe) in the list a series of rows
        allRegions[[i]] <- html_MP[prevNum:currentNum, ]
    }
}
```

## Cleaning the Dataframe
Now all the rows have been assigned to their respective dataframes in the list **allRegions**. The function *clean_Regions* takes a dataframe and returns a tidy dataframe with 4 columns. Below are some of the actions performed on each argument passed to the function argument:

1. Select the value in the first row, first column as that is the Region Name and will be used to create a new column.
2. Tidy the selected value. 
3. Remove the 4th to 6th columns.
4. Rename the columns of the dataframe.
5. Remove a few more rows, first 2 and the last row of each dataframe.
6. Assign the value from 1. to a new column **Region**
7. Reset the rows of the columns counting from 1 to the length of the dataframe.

```{r message=FALSE, warning=FALSE}
clean_Regions <- function(dframeName){
    
    # extract the region's name 
    col_Region <- dframeName[1,1]
    col_Region <- gsub("\\-.+\\]", "", col_Region)
    col_Region <- gsub("Region", "", col_Region)
    col_Region <- trimws(col_Region, "right")
    
    # delete the 4th to 6th column and update the column names
    dframeName <- dframeName[, -c(4:6)]
    colnames(dframeName) <- c("Constituency", "Elected MP", "Elected Party")
    
    # delete the 1st, 2nd and final rows
    # create a new column Region
    dframeName <- dframeName[-c(1:2, nrow(dframeName)), ]
    dframeName$Region <- col_Region
    
    rownames(dframeName) <- seq(length.out=nrow(dframeName))
    
    return(dframeName)
    
}
```

### Applying the function
There are 10 dataframes in the list *allRegions*. We can apply the function *clean_Regions* on each dataframe in the list. Using the for loop we clean each dataframe and assign it back to the list
```{r message=FALSE, warning=FALSE}
for(i in 1:length(allRegions)){
    allRegions[[i]] <- clean_Regions(allRegions[[i]])
}
```
### Bind Rows
To get all 10 cleaned dataframes into a single dataframe, we use the `bind_rows()` from the **tidyverse** package
```{r message=FALSE, warning=FALSE}
allRegions <- bind_rows(allRegions)
head(allRegions)
```

### Sanity Check
We need to perform a sanity check on the data. 
```{r message=FALSE, warning=FALSE}
nrow(allRegions)
```
The number of rows should have been 200 but it is currently 200. Checks show that the final row in the **Western Region** was removed hence it will be added and then we confirm the number of rows.
```{r message=FALSE, warning=FALSE}
allRegions <- rbind(allRegions, c("Tarkwa -Nsuaem","Mathew Kojo Kum","NDC","Western"))
dim(allRegions)
```

## Conclusion
It was a very long post, I know but I am glad you were able to come this far. 
My main take-aways are

* **rvest** is an amazing package. The *html_table()* comes in handy when one wants the data in a tabular format, remember to pass *TRUE* to the  *fill* argument.
* If a base R function can do the work, go ahead: *gsub* and *grepl* came to the rescue when I needed them.
* Use loops, use loops, use loops
* If you need to perform a set of actions more than once,put them into a function.

The final data is available [here](https://github.com/maryjonah/maryjonah.github.io/blob/master/ghana_elections/1992/parliamentary_results_1992.csv).I hope to see you in Episode 3.